Refinements for open automata

### 2021-07-30 2021-Refinement/paper/refinementarticle.tex

\begin{document}
%
\title{Refinements for open pNets}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{
Rabéa\inst{1}\orcidID{1111-2222-3333-4444} \and
Quentin\inst{1,2}\orcidID{1111-2222-3333-4444} \and
Ludo\inst{2}\orcidID{0000-1111-2222-3333} \and
Eric\inst{3}\orcidID{2222--3333-4444-5555}}
%
\authorrunning{Rab\'ea Ameur-Boulifa et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{blabla
\email{lncs@fff}\\
\url{http://www.springer.com/gp/computer-science/lncs} \and
blibli\\
\email{\{abc,def\}@fff}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
\TODO{ The abstract should briefly summarize the contents of the paper in 150--250 words.}

@@

One important aspect to notice is that in this definition, because of the symbolic nature of pNets and their structure, the fact that a guard is true is sufficient to reason on the possible paths. This slightly simplifies the definition and makes the characterisation of transition that can be triggered more symbolic. An equivalent  definition of deadlock reduction can be expressed: it states that if there is a transition that can be triggered in the first automaton, then there is a transition from the related state in the second automaton that can be triggered. This second definition is more complex, in particular because of the multiple quantifiers over transitions and automaton state.

@@

We define here simulation between labelled transition systems to Hole-equal simulation  between  open automata that have exactly the same holes.
\begin{definition}[Hole-equal simulation]
Consider two OAs \(\OAg[1]\) and \(\OAg[2]\) with  \(J_1 = J_2\), the relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is a hole-\(\symb{equal}\) simulation from $S_1$ to $S_2$ if the following conditions hold: 
\item[(1)] \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\)
\item[(2)] \(\forall \mpar{s_1, s_2} \in S_1 \times S_2,\)\vspace{-8pt}
\noindent\begin{multline*}
\mpar{\everymath{\displaystyle}\begin{array}{l}
		\bigsymb{\forall} t_1 = \OTx{1}{}{1}{1} \in \fOT{s_1}, \bigsymb{\exists} \mpar{t_{2x} = \OTx{2}{x}{2x}{2x} \in \fOT{s_2}}^{x \in X}, \\
		\quad \mpar{\forall x \in X, J'_{2x}= J'_1} \\[-10pt]
		\nwedge V_1 \uplus V_2 \uplus \fvars{t_1} \vdash \left( R\mpar{s_1, s_2} \wedge g_1 \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\begin{array}{l}
			\alpha_1 = \alpha_{2x} \wedge \bigwedge_\subbox{j \in J'_{2x}} \beta_{1j} = \beta_{2xj} \\[12pt]
			\nwedge g_{2x} \wedge R\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}}
		\end{array}}\right)
	\end{array}} \\
	\wedge \mpar{V_1 \uplus V_2 \uplus \biguplus_\subbox{t_2 \in \fOT{s_2}} \fvars{t_2} \vdash R\mpar{s_1, s_2} \wedge \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2} \implies \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1}}
\end{multline*}
\end{definition}
Note in this definition, instead of matching an instantiated transition of the first automata to another instantiated transition of the second, it matches an open transition $t_1$ to a family of covering open transitions $t_{2x}^{x\in X}$. Note also the second part of the conjunction of condition (2) expresses the deadlock reduction between automata (as introduced in Definition \ref{def:dpwd}).

Intuitively, this means that for every pair of related
states $(s_1,s_2)$  of the two automata, and for every  transition of the first automaton from $s_1$, there is a set of matching transitions  of the second automaton  from $s_2$ such that the produced action match, the actions of the same holes and the successors are related after variable update. Our definition captures a simple kind of sub-classing of open automata with the same holes. It is stronger than a strict simulation since it matches a transition with a family of transitions. With such a relation we are able to check the refinement between two open automata corresponding to the same component but specified differently, for example, by duplicating states, removing transitions,  reinforcing  guards, modifying variables, etc...

@@

The major challenge in this case is to maintain a form of transitivity while being able to take into account the actions of some of the holes. A naive definition of refinement would ensure that the holes that are identical in the two OAs are taken into account in the simulation. Unfortunately this does not work well with transitivity as if $A_1$ simulates $A_2$ and $A_2$ simulates $A_3$, and one hole appears in $A_3$ and in $A_1$ but not in $A_2$ then we have no proven property on the way $A_1$ and $A_3$ takes his hole into account, and in general  $A_1$ does not simulate $A_3$. The way we solve this issue is to remember in the simulation relation which holes have been compared. This way the relation is parameterized by the set of holes that belong to the two automata and are taken into account.
In practice, when performing successive refinements, we  keep track of which holes are in common at each step of refinement.
\begin{definition}[Open automata refinement]
For two open automata \(A_1 \defobject \OA{S_1}{s_{01}}{J_1}{V_1}{\sigma_{01}}{T_1}\) and \(A_2 \defobject \OA{S_2}{s_{02}}{J_2}{V_2}{\sigma_{02}}{T_2}\), \(A_1\) is a refinement of \(A_2\) tracking holes \(H\), noted \(\wrel{A_1}{A_2}{H}\), with \(H \subseteq J_1 \cap J_2\), is defined as:
\begin{multline*}
	\exists R: \mpar{S_1 \times S_2} \to \rformulas[V_1 \uplus V_2], \mpar{\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}} \wedge \forall \mpar{s_1, s_2} \in S_1 \times S_2, \\
	\mpar{\everymath{\displaystyle}\begin{array}{l}
		\bigsymb{\forall} \OTx{1}{}{1}{1} \in \fOT{s_1}, \bigsymb{\exists} \mpar{\OTx{2}{x}{2x}{2x} \in \fOT{s_2}}^{x \in X}, \\[12pt]
		\quad \mpar{\forall x \in X, J'_{2x} \cap H = J'_1 \cap H} \\[-10pt]
		\nwedge V_1 \uplus V_2 \uplus \fvars{t_1} \vdash R\mpar{s_1, s_2} \wedge g_1 \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\begin{array}{l}
			\alpha_1 = \alpha_{2x} \wedge \bigwedge_\subbox{j \in J'_{2x} \cap H} \beta_{1j} = \beta_{2xj} \\
			\nwedge g_{2x} \wedge R\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}}
		\end{array}} \\
	\end{array}} \\
	\wedge V_1 \uplus V_2 \uplus \biguplus_\subbox{t_2 \in \fOT{s_2}} \fvars{t_2} \vdash R\mpar{s_1, s_2} \wedge \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2} \implies \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1}
\end{multline*}
Giving \(R\) and \(H\) is sufficient to characterise the refinement, so we call \(\mpar{R, H}\) a hole-tracking simulation of \(A_1\) by \(A_2\).
Hole in \(H\) are called tracked holes.
\end{definition}
It is important to note that every action of the holes outside \(H\) is unconstrained in the related automata.
It is easy to see that the hole-equal simulation is a particular case of open automata refinement, when  $J_1=J_2=H$.

>>>>

\begin{document}
%
\title{Refinements for open automata}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{
Rabéa\inst{1}\orcidID{1111-2222-3333-4444} \and
Quentin\inst{1,2}\orcidID{1111-2222-3333-4444} \and
Ludo\inst{2}\orcidID{0000-1111-2222-3333} \and
Eric\inst{3}\orcidID{2222--3333-4444-5555}}
%
\authorrunning{Rab\'ea Ameur-Boulifa et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{blabla
\email{lncs@fff}\\
\url{http://www.springer.com/gp/computer-science/lncs} \and
blibli\\
\email{\{abc,def\}@fff}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
\TODO{ The abstract should briefly summarize the contents of the paper in 150--250 words.}

@@

One important aspect to notice is that in this definition, because of the symbolic nature of OAs and their structure, the fact that a guard is true is sufficient to reason on the possible paths. This slightly simplifies the definition and makes the characterisation of transition that can be triggered more symbolic. An equivalent  definition of deadlock reduction can be expressed: it states that if there is a transition that can be triggered in the first automaton, then there is a transition from the related state in the second automaton that can be triggered. This second definition is more complex, in particular because of the multiple quantifiers over transitions and automaton state.

@@

We define here simulation between labelled transition systems to Hole-equal simulation  between  open automata that have exactly the same holes.
\begin{definition}[Hole-equal simulation]
Consider two OAs \(\OAg[1]\) and \(\OAg[2]\) with  \(J_1 = J_2\), the relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is a hole-\(\symb{equal}\) simulation from $S_1$ to $S_2$ if the following conditions hold: 
\item[(1)] \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\)
\item[(2)] \(\forall \mpar{s_1, s_2} \in S_1 \times S_2,\)\vspace{-8pt}
\noindent\begin{multline*}
\everymath{\displaystyle}\begin{array}{l}
		\bigsymb{\forall} t_1 = \OTx{1}{}{1}{1} \in \fOT{s_1}, \bigsymb{\exists} \mpar{t_{2x} = \OTx{2}{x}{2x}{2x} \in \fOT{s_2}}^{x \in X}, \\
		\quad \mpar{\forall x \in X, J'_{2x}= J'_1} \\[-2pt]
		\nwedge V_1 \uplus V_2 \uplus \fvars{t_1} \vdash \left( R\mpar{s_1, s_2} \wedge g_1 \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\begin{array}{l}
			\alpha_1 = \alpha_{2x} \wedge \bigwedge_\subbox{j \in J'_{2x}} \beta_{1j} = \beta_{2xj} \\[12pt]
			\nwedge g_{2x} \wedge R\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}}
		\end{array}}\right)
	\end{array} 
%\\
%	\wedge \mpar{V_1 \uplus V_2 \uplus \biguplus_\subbox{t_2 \in \fOT{s_2}} \fvars{t_2} \vdash R\mpar{s_1, s_2} \wedge \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2} \implies \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1}}
\end{multline*}
\item[(3)] $R$ is deadlock reducing.
\end{definition}
Note in this definition, instead of matching an instantiated transition of the first automata to another instantiated transition of the second, it matches an open transition $t_1$ to a family of covering open transitions $t_{2x}^{x\in X}$. Note also the second part of the conjunction of condition (2) expresses the deadlock reduction between automata (as introduced in Definition \ref{def:dpwd}).

Intuitively, this means that for every pair of related
states $(s_1,s_2)$  of the two automata, and for every  transition of the first automaton from $s_1$, there is a set of matching transitions  of the second automaton  from $s_2$ such that the produced action match, the actions of the same holes and the successors are related after variable update. Our definition captures a simple kind of sub-classing of open automata with the same holes. It is stronger than a strict simulation since it matches a transition with a family of transitions. 
With such a relation we are able to check the refinement between two open automata with the same level of abstraction but specified differently, for example, by duplicating states, removing transitions,  reinforcing  guards, modifying variables. 
However, this refinement is inappropriate in the setting of composition which is the main advantage of the open automaton-based approach. 
\TODO{introduce somewhere refinement through composition} 
Refinement through composition  leads to a model in which composition can be used both to add new holes to a system and to fill holes, obtaining a system with less holes.
Next section will identify a refinement relation with the possibility to have different holes in the compared automata.

@@

The major challenge in this case is to maintain a form of transitivity while being able to take into account the actions of some of the holes. A naive definition of refinement would ensure that the holes that are identical in the two OAs are taken into account in the simulation. Unfortunately this does not work well with transitivity as if $A_1$ simulates $A_2$ and $A_2$ simulates $A_3$, and one hole appears in $A_3$ and in $A_1$ but not in $A_2$ then we have no proven property on the way $A_1$ and $A_3$ takes his hole into account, and in general  $A_1$ does not simulate $A_3$. The way we solve this issue is to remember in the simulation relation which holes have been compared. This way the relation is parameterized by the set of holes that belong to the two automata and are taken into account.
In practice, when performing successive refinements, we  keep track of which holes are in common at each step of refinement.
\begin{definition}[Open automata refinement]
For two open automata \(A_1 \defobject \OA{S_1}{s_{01}}{J_1}{V_1}{\sigma_{01}}{T_1}\) and \(A_2 \defobject \OA{S_2}{s_{02}}{J_2}{V_2}{\sigma_{02}}{T_2}\), \(A_1\) is a refinement of \(A_2\) tracking holes \(H\), noted \(\wrel{A_1}{A_2}{H}\), with \(H \subseteq J_1 \cap J_2\), if there is a deadlock reducing relation $R: \mpar{S_1 \times S_2} \to \rformulas[V_1 \uplus V_2]$ such that:
 \[\mpar{\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}}\]
and
\begin{multline*}
 \forall \mpar{s_1, s_2} \in S_1 \times S_2, \\
	\everymath{\displaystyle}\begin{array}{l}
		\bigsymb{\forall} \OTx{1}{}{1}{1} \in \fOT{s_1}, \bigsymb{\exists} \mpar{\OTx{2}{x}{2x}{2x} \in \fOT{s_2}}^{x \in X}, \\[12pt]
		\quad \mpar{\forall x \in X, J'_{2x} \cap H = J'_1 \cap H} \\[-1pt]
		\nwedge V_1 \uplus V_2 \uplus \fvars{t_1} \vdash \mpar{R\mpar{s_1, s_2} \wedge g_1 \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\begin{array}{l}
			\alpha_1 = \alpha_{2x} \wedge \bigwedge_\subbox{j \in J'_{2x} \cap H} \beta_{1j} = \beta_{2xj} \\
			\nwedge g_{2x} \wedge R\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}}
		\end{array}}} \\
	\end{array} \\
\end{multline*}

Giving \(R\) and \(H\) is sufficient to characterise the refinement, so we call \(\mpar{R, H}\) a hole-tracking simulation of \(A_1\) by \(A_2\).
Hole in \(H\) are called tracked holes.
\end{definition}
It is important to note that every action of the holes outside \(H\) is unconstrained in the related automata.
It is easy to see that the hole-equal simulation is a particular case of open automata refinement, when  $J_1=J_2=H$.
